MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-03 rebuttal.txt

Q: The statistics are described, but how would a user interpret them/make use of them for themselves? What is this shared associative structure that the model is updating? Maybe a diagram would be helpful for visualization?
A: We added a new section Statistics output and visualization, which includes how the statistics output is represented in the code and how a user can interpret and visualize it.

Q: How does the code actually deal with circular steady states?
A: If a model contains circular steady states, simulated trajectories that would indefinitely transition over circular states will eventually reach the maximum time and their simulation stops. Their statistics are further computed without any change. The original paper (Stoll et al.) discusses the ways how to detect circular steady states from the computed statistics, but we think that describing these methods is out of the scope of this paper. Therefore, we included general information about circular steady states in Statistics output and visualization section and referenced the original paper for the reader who wants to know more detailed information.

Q: There needs to be a description about the original code. Most importantly, it is always referred to, as the CPU version. This suggests, that we are talking about a purely sequential code. In the Methodology, however, it is described to be run on a full 32-core node, which implies already some basic parallelization. This should be made clear from the beginning.
A: Indeed, the implementation of the "CPU version" was early on parallelized to work on multiple cores, using POSIX threads. We added a new subsection to the background section (MaBoSS CPU Implementation) to describe this implementation.

Q: The Boolean signaling modeling paragraph needs to be extended in more details: To me it is not clear, why e.g., the transition rates should be in [0, inf) and not [0,1], since we are talking about probabilities. The concept of asynchronous updating should be made a bit clearer as well. Furthermore, it is often referred to the evaluation of the Boolean formulas as the most expensive computational part. It would help to get an idea, how such a formula would look like, and how they are implemented, to give the reader context about this problem. Furthermore, the word "node" is used quite often in the paper, so it should be made clear, what this refers to exactly.
A: We extended the background section with all your comments. Regarding the node, [0, inf) transition rates and asynchronous updates, we added explanations to the first paragraphs of sections 2.1 and 2.2. In order to add an example of Boolean formula, we explained it in more detail by defining its binary function, which determines the transition rate using a simple rule in equation 5. Having the binary function, we presented its simple example in equation 4. The way how the boolean formulas are implemented in the code is described in the first paragraph of section 3.1, simply referring to it as a recursive traversal of an expression tree.

Q: Algorithm 1 should be made clearer: how are transition rates computed? What does the bit-flip mean? delta t is computed using a "u" - where does that come from?
A: Transition rates are computed by evaluating boolean formulas of each node, as is written on line 2 of Algorithm 1. To make it clearer, we modified the second paragraph of section 2.2, where we explicitly state, that the output of boolean logic is the transition rates. Regarding "u", the Gillespie algorithm uses 2 random numbers (r,u) drawn from the uniform distribution to randomly select the transition node and the transition time. We added comments to the algorithm and added a line referencing that the algorithm follows the Gillespie algorithm.

@@

Q: Since the authors have disabled issue tracking on their github repositories, I urge them to add a contact method to sysbio-curie/hpcmaboss-artifact and sysbio-curie/MaBoSS.GPU, for users to obtain assistance and provide feedback, including bug reports.
A: This was a mistake, we now have enabled issue tracking on MaBoSS.GPU and hpcmaboss-artifact repositories. Thanks for reminding us.

Q:  It is not clear whether the version provided via the colomoto conda repository is MPI enabled, so a clarification is needed.
A: This is something we have been considering, but we have not moved forward yet. We will most likely have to create specific conda packages which are MPI enabled, in addition to the classic ones. Same thing for MaBoSS.GPU. We also need to think about this for the colomoto docker, but again it might not be straightforward. For now, we think that the target users of these new implementations are knowledgeable enough to install them manually.

Q: A concern is that the provided background is insufficient for the applicability and usefulness of the software to be apparent to the broader audience of BMC Bioinformatics. The authors should add more to provide context and insight into the interpretation and use of MaBoSS output, for those not already familiar with the previous MaBoSS versions.
A: We included a new section Statistics output and visualization.

Q: Finally, to improve accessibility, the following could do with explanation: variables “u” and “δt” in Algorithm 1, “PRAM”, “CNF”, “DNF”.
A: We explained the acronyms and provided comments for “δt” in the algorithm. “u” is to provide randomness in “δt” selection, which is now highlighted in the comment. Moreover, we reference the Gillespie algorithm, on which our algorithm is based.

### 2024-05-03 Article.tex
finish background

A Boolean signaling model consists of $n$ \emph{nodes} that are either active or inactive, gaining binary values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ binary values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

@@

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where at most a single node changes its value in each transition. Therefore, there is a transition from $S$ to $S'$ only if it holds that 
\begin{equation}
    \begin{split}
S_j \ne S'_j & \text{ for a given } j \\
S_i = S'_i & \text{ for } i \ne j.
    \end{split}
\end{equation}
Consequently, $S$ can have at most $n$ possible transitions. In programming terms, $S'$ is obtained by flipping the $j$-th bit of $S$.

@@

The main computational part of a Boolean logic is a \emph{Boolean formula} $B: \mathcal{S} \rightarrow \{0, 1\}$, which consists of logical operators (such as \emph{and}, \emph{or}, \emph{xor}, \emph{not}) with nodes as operands. For example, 
\begin{equation}
    B_i(S) = (S_2 \land S_3) \lor S_4 
\end{equation}
is a Boolean formula for node $i$, where $S_2$, $S_3$ and $S_4$ are binary values of nodes $2$, $3$ and $4$, respectively. Using the values of $B(S)$ and $S_i$, $\mathcal{B}_i(S)$ is defined by a simple rule:
\begin{equation}
    \mathcal{B}_i(S) = \begin{cases}
        0 & \text{if } S_i = B_i(S) \\
        r & \text{otherwise}
    \end{cases}
    \label{eq:transition_rate}
\end{equation}

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a \emph{fixed point} state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}.

\begin{algorithm}
\caption{A single iteration of the MaBoSS simulation of a trajectory, given the state $S$ and time $t$.}
\label{alg:iter}
\begin{algorithmic}[1]
\Procedure{TrajectorySimulationStep}{S, t}
\State $\rho_1 \gets \mathcal{B}_1(S), \dots, \rho_n \gets \mathcal{B}_n(S)$ \Comment compute transition rates
\State $r \gets \text{random}([0, \sum_{i=1}^n \rho_i))$
\State $i \gets \min_i \sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$ \Comment select index of the transition node
\State $S'\gets S\ \text{with the $i$-th bit flipped}$
\State $u \gets \text{random}([0, 1])$
\State $\delta t \gets -\frac{\ln u}{\sum_{i=1}^n \rho_i}$ \Comment compute time to the next transition
\State \textbf{return} $(S', t + \delta t)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

To obtain a good overview of the Markovian process, we apply the Gillespie stochastic simulation algorithm~\cite{gillespie1976general} (or kinetic Monte-Carlo~\cite{young1966monte}) on the Boolean state space. Multiple trajectories are generated and aggregated in compound trajectory statistics. Commonly obtained statistics include:
\begin{itemize}
    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size. For each window, the probability of each state is computed as the duration spent in the state divided by the window size. The probabilities of the corresponding windows are then averaged across all subtrajectories.
    \item \emph{Final states} --- The last sampled states from the trajectories are used to compute a final state distribution.
    \item \emph{Fixed states} --- All reached fixed points are used to compute a fixed state distribution.
\end{itemize}

@@

\begin{table}
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 10\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{The main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of present nodes in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}

@@

>>>>

A Boolean signaling model consists of $n$ \emph{nodes}, which can represent a gene, protein or an event in a cell. Nodes are either active or inactive, gaining binary values 1 or 0 respectively. The \emph{state} of the whole model is represented by a vector $S$ of $n$ binary values where $S_i$ represents the value of the $i$-th node. We denote the set of all possible states as $\mathcal{S} = \{0, 1\}^n$; thus $|\mathcal{S}| = 2^n$.

@@

\subsection{MaBoSS: Markovian Boolean Stochastic Simulator}
MaBoSS simulates the \emph{asynchronous update strategy}, where only a single node changes its value in each transition (contrary to the \emph{synchronous update strategy}, for which all nodes that can be updated are updated~\cite{stoll2012continuous}). Therefore, there is a transition from $S$ to $S'$ only if it holds that 
\begin{equation}
    \begin{split}
S_j \ne S'_j & \text{ for a given } j \\
S_i = S'_i & \text{ for } i \ne j.
    \end{split}
\end{equation}
Consequently, $S$ can have at most $n$ possible transitions. In programming terms, $S'$ is obtained by flipping the $j$-th bit of $S$.

@@

The main computational part of the Boolean logic is its binary function $f: \mathcal{S} \rightarrow \{0, 1\}$, which consists of logical operators (such as \emph{and}, \emph{or}, \emph{xor}, \emph{not}) with nodes as operands. For example, 
\begin{equation}
    f_i(S) = (S_2 \land S_3) \lor S_4 
\end{equation}
is the binary function for node $i$, having nodes $2$, $3$ and $4$ as its operands.
The binary function of a node determines the value to which the node can transition. Thus, $S$ can transition at node $i$ at rate $r$ only if $f_i(S) \neq S_i$. Concisely, $\mathcal{B}_i$ is defined as\footnote{Generally, $\mathcal{B}_i$ can be defined using a pair of binary formulas~\cite{stoll2012continuous}: one used when a node is active and one when it is inactive. This results in a slightly more branched definition, which we omitted for brevity.}
\begin{equation}
    \mathcal{B}_i(S) = \begin{cases}
        0 & \text{if } S_i = f_i(S) \\
        r & \text{otherwise}
    \end{cases}
    \label{eq:transition_rate}
\end{equation}

MaBoSS algorithm simulates the above process to produce stochastic \emph{trajectories}: sequences of states $S^0, S^1, \dots, S^k$ and time points $t^0 < t^1 < \dots < t^k$ where $t^0 = 0$ and $S^0$ is the initial state, and for each $i \in \{0, \dots, k-1\}$, $S^i$ transitions to $S^{i+1}$ at time $t^{i+1}$. The simulation ends either by a timeout when reaching the maximal allowed time, or by reaching a \emph{fixed point} state with no outgoing transitions. The algorithm for a single iteration of the trajectory simulation is given explicitly in Algorithm~\ref{alg:iter}, which is the direct application of the Gillespie stochastic simulation algorithm~\cite{gillespie1976general} (or kinetic Monte-Carlo~\cite{young1966monte}) on the Boolean state space.

\begin{algorithm}
\caption{A single iteration of the MaBoSS simulation of a trajectory, given the state $S$ and time $t$.}
\label{alg:iter}
\begin{algorithmic}[1]
\Procedure{TrajectorySimulationStep}{S, t}
\State $\rho_1 \gets \mathcal{B}_1(S), \dots, \rho_n \gets \mathcal{B}_n(S)$ \Comment compute transition rates
\State $r \gets \text{random}([0, \sum_{i=1}^n \rho_i))$
\State $i \gets \min_i \sum_{j=1}^{i-1} \rho_j \leq r < \sum_{j=1}^{i} \rho_j$ \Comment randomly select the transition node
\State $S'\gets S\ \text{with the $i$-th bit flipped}$
\State $u \gets \text{random}([0, 1])$
\State $\delta t \gets -\frac{\ln u}{\sum_{i=1}^n \rho_i}$ \Comment randomly select time to the next transition
\State \textbf{return} $(S', t + \delta t)$
\EndProcedure
\end{algorithmic}
\end{algorithm}

Multiple trajectories are generated and aggregated in compound trajectory statistics. Commonly obtained statistics include:
\begin{itemize}
    \item \emph{Network state probabilities on a time window} --- Trajectory states are divided by their transition times into time windows based on the time intervals specified by a window size. For each window, the probability of each state is computed as the duration spent in the state divided by the window size. The probabilities of the corresponding windows are then averaged across all subtrajectories.
    \item \emph{Final states} --- The last sampled states from the trajectories are used to compute a final state distribution.
    \item \emph{Fixed states} --- All reached fixed points are used to compute a fixed state distribution.
\end{itemize}

@@

@@

\begin{table}[h]
    \begin{tabular}{l|cccc}
        \textbf{Model}     & \textbf{\# Nodes (non-inter.)} & \textbf{\# Traj.}        & \textbf{Avg. formula size} & \textbf{Avg. traj. length} \\
        \midrule
        \textsc{cellcycle} & $10$ $(4)$                     & $1$M                     & $4$                        & $26$                       \\
        \textsc{sizek}     & $87$ $(4)$                     & $1$M                     & $22$                       & $525$                      \\
        \textsc{Montagud}  & $133$ $(3)$                    & $1$M                     & $4$                        & $197$                      \\
        \hline
        \emph{Synthetic}   & $10 - 1000$ $(5)$              & $1\text{M} - 10\text{M}$ & $10 - 100$                 & $100$                      \\
        \bottomrule
    \end{tabular}
    \centering
    \caption{The main features of the synthetic and real-world models used in the benchmarks. It includes the size of models in terms of nodes and non-internal nodes, the number of simulated trajectories, the average formula size measured as the arithmetic mean of operands count in each formula, and the average length of all simulated trajectories. Note that not all combinations of features for the synthetic model were used in the benchmarks, see the following figures for more details.}
    \label{tab:model_features}
\end{table}
